Boson-Sampling with Single Photons: A Roadmap to Post-Classical Computation

### 2016-09-21 paper.tex
Completed body of text - awaiting appendix from Zuen

The brightness of $n$ photons will decay exponentially as $p^n$ if each SPS has efficiency $p$. For current mature heralded SPSs based on spontaneous parametric down-conversion (SPDC), one solution is to introduce more SPDC sources and make $p$ approach unity by the technique of multiplexing \cite{bib:34}. For boson-sampling, there is an even simpler method known as `scattershot' boson-sampling (SBS) \cite{bib:35, bib:36}, which allows us to perform an identically hard computation without deterministic SPSs or multiplexers. The core idea is to run $m$ SPDCs simultaneously, and post-select the instances where $n$ heralded detectors are triggered, regardless of which detectors click. SBS seems something of a `double sampling' problem, which can be decomposed into uniform sampling at the input state, followed by ordinary boson-sampling. But each single sample is an instance of original boson-sampling with different input states. In fact, Aaronson \cite{bib:35} and Lund \cite{bib:36} pointed out that SBS remains in the same computational complexity class. The first 3-photon experiment has been implemented with 6 equivalent SPDC sources in a 13-mode circuit \cite{bib:38}. The model is shown in Fig.~\ref{fig:scattershot_model}.

\begin{figure}[!htb]
\includegraphics[width=\columnwidth]{scattershot_model}
\caption{The scheme for scattershot boson-sampling \cite{bib:36}. For an \mbox{$m\times m$}-mode interferometer, there are $m$ pairs of SPDCs, of which only $n$ generate single photons at a time.} \label{fig:scattershot_model}
\end{figure}

It is well-known that the two-mode squeezed state via SPDC has the form
\begin{align}
\ket\psi_\mathrm{SPDC} = \sqrt{1-\chi^2}\sum_{N=0}^\infty \ket{N,N},
\end{align}
where $N$ is photon-number and $\chi$ is the squeezing parameter, related to the nonlinearity of the crystal, the pump amplitude and the crystal length. Thus the SPS efficiency is $p=\chi^2(1-\chi^2)$. The model of SBS is illustrated in Fig.~\ref{fig:scattershot_model}. There are in total $\binom{m}{n}$ combinations of different acceptable input configurations, and the generation probability of $n$ single photons is therefore promoted to,
\begin{align}
P(n) = \binom{m}{n} \chi^{2n}(1-\chi^2)^{n^2}.
\end{align}

@@

As pointed out in Sec.~\ref{sec:sampling_time}, the only thing we care about in the competition of quantum boson-samplers versus classical simulators is who would be the first to produce a legitimate sample from the permanent-related probability distribution. And in the context of no a priori knowledge about which input configuration would be picked in the experiment, to simulate SBS, a classical computer can obtain a legitimate sample just by simulating ordinary boson-sampling with deterministic input states. Thus, we follow the same $t_c$ in Eq.~\ref{eq:tau_c} for SBS, which does not need to be multiplied by a combinatorial factor at all as in \cite{bib:37}.

By comparing $\tau_q$ and $\tau_c$, in Fig.~\ref{fig:curve_scattershot} we plot the supremacy boundaries for SBS using nondeterministic SPS via SPDC. We can see that, with the help of the number of different combinations, SPDC survives and SBS is a feasible approach to beating advanced classical competitors. However, in order to scale SBS to the interesting regime of $20\sim 30$ photons, other challenges remain challenging. First, SBS requires $n^2$ SPDC sources, quadratically more than the usual $n$. And a larger-depth optical network is required to fully implement an $m\times m$ unitary matrix, rather than in original boson-sampling, with a fixed number of $n$ input modes, it can be simplified to an $m\times n$ matrix. Besides, as twice the number of photodetectors will be employed, to assure a certain sample rate, the requirement of detection efficiency imposed on photodetectors of SBS would be higher. Comparing Fig.~\ref{curves_1} and Fig.~\ref{fig:curve_scattershot}, we see that, to beat a classical computer, SBS should either scale to a large size or require the quantum device to have a higher total efficiency with respect to ordinary boson-sampling.

@@

@@

The relationship between the single-photon input and net output fidelities are illustrated in Fig.~\ref{fig:fidelity}. Based on this, to achieve an output fidelity of \mbox{$F_\mathrm{out} = 0.9$} (i.e 90\% confidence that a given sample is a legitimate boson-sampling sample) requires single-photon input fidelities of \mbox{$F_\mathrm{in} = 0.995$} for 10 photons, and \mbox{$F_\mathrm{in} = 0.998$} for 20 photons. This type of single-photon interference fidelity is realistic using present-day technology, which comes very close to unity.

>>>>

The brightness of $n$ photons will decay exponentially as $p^n$ if each SPS has efficiency $p$. For current mature heralded SPSs based on spontaneous parametric down-conversion (SPDC), one solution is to introduce more SPDC sources and make $p$ approach unity by the technique of multiplexing \cite{bib:34}. The idea is simple -- we prepare a large bank of SPDC sources triggering in parallel, all feeding into a multiplexer, which routes successfully heralded photons to the desired outputs. This allows state preparation arbitrarily close to unity in principle (in the ideal lossless case), with a sufficient number of sources operating in parallel.

`Scattershot' boson-sampling (SBS) is a much simpler solution to the same problem \cite{bib:35, bib:36}. It offers the major advantage that the multiplexer may be simply eliminated, thereby completely mitigating the need for technologically challenging fast switching and feedforward. The core idea is to run $m$ SPDCs simultaneously (one per input mode), and post-select the instances where $n$ heralded detectors are triggered, regardless of in which configuration the detectors click. SBS is something of a `double sampling' problem, which can be decomposed into uniform sampling at the input state, followed by ordinary boson-sampling. But each single sample is an instance of original boson-sampling with different input states. In fact, Aaronson \cite{bib:35} and Lund et al. \cite{bib:36} pointed out that SBS remains in the same computational complexity class as conventional boson-sampling. The first 3-photon SBS experiment has been implemented with 6 identical SPDC sources in a 13-mode circuit \cite{bib:38}. The model is shown in Fig.~\ref{fig:scattershot_model}.

\begin{figure}[!htb]
\includegraphics[width=\columnwidth]{scattershot_model}
\caption{The scheme for scattershot boson-sampling \cite{bib:36}. For an \mbox{$m\times m$}-mode interferometer, there are $m$ pairs of SPDCs, of which, upon post-selection, only $n$ generate single photons at a time.} \label{fig:scattershot_model}
\end{figure}

An intuitive explanation for the computational complexity of SBS is that in ordinary boson-sampling we are sampling from a distribution of permanents of Haar-random matrices. The SBS protocol simply adds an extra level of randomisation -- specifically, a random permutation -- prior to the Haar-random matrix. Intuitively, we obviously expect that a permutation of a random matrix yields another random matrix.

We now turn our attention to the efficiency of SBS. It is well-known that the two-mode squeezed state via SPDC has the form,
\begin{align}
\ket\psi_\mathrm{SPDC} = \sqrt{1-\chi^2}\sum_{N=0}^\infty \ket{N,N},
\end{align}
where $N$ is photon-number (always the same for both modes) and $\chi$ is the squeezing parameter, related to the nonlinearity of the crystal, the pump amplitude and the crystal length. Thus the SPS efficiency is \mbox{$p=\chi^2(1-\chi^2)$}. Referring to the SBS model illustrated in Fig.~\ref{fig:scattershot_model}, there are in total $\binom{m}{n}$ combinations of different acceptable input configurations, and the generation probability of $n$ single photons is therefore promoted to,
\begin{align}
P(n) = \binom{m}{n} \chi^{2n}(1-\chi^2)^{n^2}.
\end{align}

@@

As pointed out in Sec.~\ref{sec:sampling_time}, the only thing we care about in the competition of quantum boson-samplers versus classical simulators is who would be the first to produce a legitimate sample from the permanent-related probability distribution. And in the context of no a priori knowledge about which input configuration would be picked in the experiment, to simulate SBS, a classical computer can obtain a legitimate sample just by simulating ordinary boson-sampling with randomly permuted input states, which does not incur any computational overhead, i.e the `hard' part of the computation (calculating permanents) remains unchanged. Thus, we follow the same $t_c$ in Eq.~\ref{eq:tau_c} for SBS, which does not need to be multiplied by a combinatorial factor at all as in \cite{bib:37}.

By comparing $\tau_q$ and $\tau_c$, in Fig.~\ref{fig:curve_scattershot} we plot the supremacy boundaries for SBS using nondeterministic SPS via SPDC. We can see that, with the help of the number of different combinations, SPDC survives and SBS is a feasible approach to beating advanced classical competitors. However, in order to scale SBS to the interesting regime of $20\sim 30$ photons, other challenges remain challenging. First, SBS requires $n^2$ SPDC sources, quadratically more than the usual $n$. And a larger-depth optical network is required to fully implement an $m\times m$ unitary matrix, rather than in original boson-sampling, with a fixed number of $n$ input modes, it can be simplified to an $m\times n$ matrix. Besides, as twice the number of photodetectors will be employed, to assure a certain sample rate, the requirement of detection efficiency imposed on photodetectors of SBS would be higher. Comparing Fig.~\ref{fig:curves_1} and Fig.~\ref{fig:curve_scattershot}, we see that, to beat a classical computer, SBS should either scale to a large size or require the quantum device to have a higher total efficiency with respect to ordinary boson-sampling.

@@

\subsubsection{Analysis}

The scattershot method is a recipe for overcoming the exponentially decaying efficiency of parallel SPDC sources. Nonetheless, despite this improvement, there is still a major problem to consider -- efficiency. Here, we discuss the two dominant factors -- coupling and detection efficiency.

On one hand, the problem results from the large number ($m$ in total) of SPDC sources that SBS needs. As we mentioned before, we will record the instance when both $n$ heralding detectors and $n$ detectors at the output modes are fired. However, single photon pairs might be generated from more than $n$ SPDC sources and coupled into the interferometer without being consciously noticed by us (without our awareness), as long as the extra photons were not successfully heralded due to the detection inefficiency. If an unheralded photon is detected in the output, but a correct heralded photon is lost in the interferometer, a wrong sample will be taken. Enhancing the detection efficiency can relax the problem, whereas to put an end to `dark photons' we need high speed vacuum shutters to close the input modes whose corresponding triggering detectors did not click \cite{bib:38}.

On the other hand, SPDC sources exhibit higher-order emission, leading to the possibility that multiple photon pairs might be generated. The following two approaches may be helpful: attenuating the brightness of the source by decreasing the pump power of the SPDCs, or employing photon-number-resolving detectors \cite{bib:34}.

When the detectors are perfect but non-number-resolving, the probability that $n$ heralding detectors are triggered, and all of them result from single photon pairs generated by $n$ SPDCs (ignoring higher order terms) is,
\begin{align}
p_\mathrm{input-correct} = \binom{m}{n} p^n(1-p-p^2)^{m-n}.
\end{align}
The total probability that $n$ SPDC sources generate photons is given by
\begin{align}
p_\mathrm{input-total} = \sum_{i=0}^n \binom{m}{n-i,i}p^{n+i}(1-p-p^2)^{m-n}.
\end{align}
\comment{What's the funny binomial?}. Nevertheless, we cannot distinguish the correct preparation of $n$ single photons from the total occasions of $n$ heralded sources. As shown in Fig.~\ref{fig:SBS_pair_gen}, the correct generation ratio defined as $p_\mathrm{input-correct}/p_\mathrm{input-total}$ decreases rapidly with increasing $n$.

For example, for $n=6$, almost half the input states are erroneous, where some two-photon events are mixed. Without photon-number-resolving detectors, we cannot exclude this erroneous component before coupling photons into the interferometer. Consider the first possible remedy, attenuating the pump power to reducing the generation probability of higher-order terms at the cost of the brightness of the SPS. For simplicity, let us consider the example of $n=20$. To ensure the correct rate of input states is $>0.9$, we should keep $p<0.006$, much smaller than the scalable condition that $p$ equals to \mbox{$20/21^2=0.04$}, leading to the decrease of source brightness of more than 10 orders of magnitude. What does it imply? If all the coupling and detection processes are perfect, it implies we obtain 6 counts a day for 76MHz repeat frequency. Let alone beating classical simulation, it is too difficult to keep the system stable in the data collection period. Therefore, only the second remedy -- photon-number-resolving detectors -- remains probable and useful to implement scalable SBS.

We take shutters and photon number resolving detectors into consideration with different configurations. As shown in Fig.~\ref{fig:SBS_pair_gen}, only if both of them are exploited, the rigorous requirements in detection efficiency and coupling efficiency of the quantum system can be greatly relaxed. In this case, 20-photon SBS seems viable if the detection efficiency and coupling efficiency of the quantum system reach 0.95 and 0.8, respectively.

\begin{figure}[!htb]
\includegraphics[width=\columnwidth]{SBS_pair_gen}
\caption{The correct generation ratio of $n$ single photons in \mbox{$m=n^2$} pairs of SPDCs.} \label{fig:SBS_pair_gen}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\columnwidth]{SBS_eff}
\caption{The requirements of coupling and detection efficiency for scattershot boson-sampling to guarantee SNR$>$2 when increasing photon number: (a) without shutters and number-resolving detectors; (b) with shutters but without number-resolving detectors; (c) without shutters but with number-resolving detectors; (d) with shutters and number-resolving detectors. The efficiency denoted on the vertical axis is the coupling efficiency from SPDC sources to the interferometer. The numbers labeled in the broken lines indicate different detection efficiencies of the quantum system.} \label{fig:SBS_eff}
\end{figure}

@@

The relationship between the single-photon input and net output fidelities are illustrated in Fig.~\ref{fig:fidelity}. Based on this, to achieve an output fidelity of \mbox{$F_\mathrm{out} = 0.9$} (i.e 90\% confidence that a given sample is taken from a legitimate boson-sampling distribution) requires single-photon input fidelities of \mbox{$F_\mathrm{in} = 0.995$} for 10 photons, and \mbox{$F_\mathrm{in} = 0.998$} for 20 photons. This type of single-photon interference fidelity is realistic using present-day technology, which comes very close to unity.
